---
title: "Interactivity with gg2d3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactivity with gg2d3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

gg2d3 provides a composable pipe-based API for adding interactivity to D3
widgets. Each function (`d3_tooltip()`, `d3_hover()`, `d3_zoom()`,
`d3_brush()`) takes a widget and returns a widget, so they chain naturally and
every verb is optional.

## The pipe API

Call `gg2d3()` on any ggplot, then pipe the result through whichever
interactivity verbs you want. Order within the chain does not matter — the
functions are additive and non-destructive.

```{r}
library(ggplot2)
library(gg2d3)

p <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(size = 3)

# Full interactive stack — tooltip + hover + zoom + brush
gg2d3(p) |>
  d3_tooltip() |>
  d3_hover() |>
  d3_zoom() |>
  d3_brush()
```

You can use any subset:

```{r}
# Tooltip and zoom only
gg2d3(p) |> d3_tooltip() |> d3_zoom()

# Brush only
gg2d3(p) |> d3_brush()
```

## Tooltips (`d3_tooltip`)

`d3_tooltip()` shows a floating tooltip containing data values when the user
hovers over a visual element (point, bar, line segment, etc.).

**Signature:** `d3_tooltip(widget, fields = NULL, formatter = NULL)`

```{r}
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) + geom_point()

# Default: show all mapped aesthetics
gg2d3(p) |> d3_tooltip()

# Show only a subset of fields
gg2d3(p) |> d3_tooltip(fields = c("wt", "mpg"))

# Custom JavaScript formatter
# formatter receives (field, value) and returns a display string
gg2d3(p) |> d3_tooltip(
  formatter = "function(field, value) {
    if (field === 'mpg') return field + ': ' + value + ' mpg';
    return field + ': ' + value;
  }"
)
```

Tooltip behavior:

- When `fields = NULL` (default), all aesthetics except internal fields
  (those starting with underscore, plus `PANEL`, `group`, etc.) are shown.
- Date and datetime values are automatically formatted using the browser's
  locale.
- The tooltip follows the cursor and disappears on mouse-out.

## Hover highlighting (`d3_hover`)

`d3_hover()` dims non-hovered elements so the hovered point or group stands
out visually.

**Signature:** `d3_hover(widget, opacity = 0.3, stroke = NULL, stroke_width = NULL)`

```{r}
p <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(size = 3)

# Default: dim others to 30% opacity; no highlight stroke
gg2d3(p) |> d3_hover()

# Stronger dimming
gg2d3(p) |> d3_hover(opacity = 0.2)

# Add a visible outline on the hovered element
gg2d3(p) |> d3_hover(opacity = 0.3, stroke = "black", stroke_width = 2)

# Combine with tooltip
gg2d3(p) |> d3_tooltip() |> d3_hover(opacity = 0.5)
```

Hover behavior:

- The hovered element always has opacity 1.0; `opacity` controls the
  non-hovered elements only.
- `stroke` and `stroke_width` are only applied to the hovered element; if
  `stroke = NULL` (default), no outline is added.
- When a brush selection is active, hover dimming is automatically disabled to
  avoid visual conflicts. Hover resumes when the brush is cleared.

## Zoom and pan (`d3_zoom`)

`d3_zoom()` enables scroll-wheel zoom and drag-to-pan. Double-click resets
to the original view.

**Signature:** `d3_zoom(widget, scale_extent = c(1, 8), direction = c("both", "x", "y"))`

```{r}
p <- ggplot(economics, aes(date, unemploy)) + geom_line()

# Default: zoom both axes, 1x minimum to 8x maximum
gg2d3(p) |> d3_zoom()

# Zoom x-axis only (useful for time series)
gg2d3(p) |> d3_zoom(direction = "x")

# Zoom y-axis only
gg2d3(p) |> d3_zoom(direction = "y")

# Allow up to 20x zoom
gg2d3(p) |> d3_zoom(scale_extent = c(1, 20))

# x-axis only with wider zoom range
gg2d3(p) |> d3_zoom(direction = "x", scale_extent = c(1, 20))
```

Zoom behavior:

- `scale_extent[1]` must be >= 1 (no zoom out beyond the original view).
- Axes — including temporal axes and secondary axes — update dynamically
  during zoom. Date formatting is preserved.
- Drag-to-pan is constrained to the panel background so that clicking data
  elements does not trigger accidental panning.
- Double-click resets zoom and pan to the original view.

## Brush selection (`d3_brush`)

`d3_brush()` lets users drag to define a rectangular selection. Selected
elements remain at full opacity; unselected elements are dimmed. Double-click
clears the selection.

**Signature:** `d3_brush(widget, direction = c("xy", "x", "y"), on_brush = NULL, fill = "#3b82f6", opacity = 0.15)`

```{r}
p <- ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
  geom_point(size = 2)

# Default: 2D rectangular brush, blue overlay, non-selected dims to 0.15
gg2d3(p) |> d3_brush()

# Horizontal band only (select by x range)
gg2d3(p) |> d3_brush(direction = "x")

# Vertical band only (select by y range)
gg2d3(p) |> d3_brush(direction = "y")

# Custom overlay color and non-selected opacity
gg2d3(p) |> d3_brush(fill = "#ef4444", opacity = 0.3)

# JavaScript callback receiving selected data array
gg2d3(p) |> d3_brush(
  on_brush = "function(selectedData) {
    console.log(selectedData.length + ' points selected');
  }"
)

# Combine with tooltip
gg2d3(p) |> d3_tooltip() |> d3_brush()
```

Brush behavior:

- `fill` is the color of the translucent brush overlay rectangle
  (default `"#3b82f6"`, a light blue).
- `opacity` controls unselected elements; selected elements are always at
  opacity 1.0.
- `on_brush` receives the selected data rows as a JavaScript array; each
  element is the row object as passed to D3.
- Double-click anywhere in the panel clears the brush and restores all
  elements to full opacity.

## Linked views with Crosstalk

gg2d3 integrates with the
[crosstalk](https://rstudio.github.io/crosstalk/) package for linked
brushing across multiple widgets. Wrap your data frame in
`crosstalk::SharedData$new()` before building the ggplot, then give that
shared data to each widget. Brushing in one widget automatically
cross-highlights the corresponding observations in all linked widgets.

```{r}
has_crosstalk <- requireNamespace("crosstalk", quietly = TRUE)
has_htmltools <- requireNamespace("htmltools", quietly = TRUE)

if (!has_crosstalk || !has_htmltools) {
  missing_crosstalk_packages <- c(
    if (!has_crosstalk) "crosstalk",
    if (!has_htmltools) "htmltools"
  )
  cat(
    "PKGDOWN_CROSSTALK_OPTIONAL_SKIP: linked-view example not rendered; missing ",
    paste(missing_crosstalk_packages, collapse = ", "),
    ".\n",
    sep = ""
  )
} else {
  # Wrap dataset in SharedData; both plots share explicit stable keys/group.
  shared_iris <- crosstalk::SharedData$new(
    iris,
    key = rownames(iris),
    group = "pkgdown_crosstalk_iris"
  )

  p1 <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
    geom_point(size = 2)
  p1$data <- shared_iris

  p2 <- ggplot(iris, aes(Petal.Length, Petal.Width, color = Species)) +
    geom_point(size = 2)
  p2$data <- shared_iris

  w1 <- gg2d3(p1) |> d3_tooltip() |> d3_brush()
  w2 <- gg2d3(p2) |> d3_tooltip() |> d3_brush()

  htmltools::tagList(
    htmltools::div(
      style = paste(
        "display: grid;",
        "grid-template-columns: repeat(auto-fit, minmax(320px, 1fr));",
        "gap: 1rem;"
      ),
      w1,
      w2
    )
  )
}
```

Crosstalk behavior:

- **No Shiny required** — linked views work in static HTML documents (e.g.,
  knitted R Markdown or `htmltools::save_html()`).
- The `SharedData` group name defaults to a random UUID; pass
  `group = "my_group"` to `SharedData$new()` if you need to link widgets
  across separate code chunks.
- Each widget can independently carry any combination of `d3_tooltip()`,
  `d3_hover()`, `d3_zoom()`, and `d3_brush()` — Crosstalk operates at the
  data key level, not at the interactivity verb level.

## Interaction caveats

- **Brush overrides hover** — when a brush selection is active, hover dimming
  is suspended. Hover resumes as soon as the brush is cleared (double-click).
- **Zoom and brush coexist** — zooming does not clear an existing brush
  selection; the selection rectangle scales with the view.
- **Large datasets** — for datasets above ~10 k rows, prefer `d3_tooltip()`
  over `d3_hover()`. Tooltip hit-testing is O(1) per element; hover dimming
  iterates all elements on every mouse event.
- **Faceted plots** — brush selection is per-panel. Selecting in one facet
  does not dim elements in other facets; use Crosstalk for cross-panel
  linking.
- **`geom_polygon` layers** — tooltip, hover, custom handlers, Shiny-style click
  handlers, brush selection, and Crosstalk target `path.geom-polygon` marks.
  Public callback payloads are sanitized representative source-row objects.
  Brush selection uses the SVG path bounds, not sf-style projected anchor
  points or topology-aware polygon overlap.
- **`geom_sf` layers** — tooltip, hover, custom handlers, Shiny-style click
  handlers, and brush callbacks target `.geom-sf` polygon, point, and line
  marks. Public callback payloads are sanitized source-row objects with
  renderer-only fields removed. Brush selection uses rendered representative
  anchors from `data-cx` and `data-cy` rather than true geometry-overlap
  brushing; these representative anchors are stable panel-local hit-test
  points. `d3_zoom()` warns with `d3_zoom() does not support geom_sf layers yet; zoom has been suppressed.` and returns the unzoomed widget.
  Optional browser validation for these sf interactions is R/testthat/chromote
  based and may skip cleanly when optional local tooling is unavailable.
- **`geom_sf_text` and `geom_sf_label` layers** — sf annotations share the
  `.geom-sf` interaction path. `geom_sf_text()` targets
  `text.geom-sf.geom-sf-text`; `geom_sf_label()` targets
  `g.geom-sf.geom-sf-label`. Tooltip, hover, handlers, and Shiny-style clicks
  receive sanitized public payloads. Brush selection uses projected anchor
  coordinates, so label bounds are not treated as topology-aware geometry
  intersections.
- **Geometry caveats** — polygon, rect/tile, SF annotation, and map-feature
  limitations are summarized in the relevant sections of the main vignette.
- **Zoom extent** — `scale_extent[1]` must be >= 1. Setting it below 1 would
  allow zooming further out than the original view, which distorts axis
  labels; the function raises an error instead.
